A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-02-20 paper/paper.tex

Water stands out among liquids for its extraordinary properties.  Water is
the only natural substance that is found in solid, liquid and gaseous
states at temperatures and pressures normally found on Earth.  Water has a
density maximum at 4$^\circ$C, and expands upon freezing.  The heat
capacity of water is the highest of any liquid, and its latent heat of
fusion and vaporization are also the highest.  Water's thermal conductivity
and surface tensions are the highest of any non-metallic liquid.  Water has
an unusually high dielectric constant.
Much of the unusual behavior of water is explained by the ability of each
water molecule to form \emph{four} strong, directional hydrogen bonds---two
acceptors and two donors.  This directional bonding dictates the open
structure of ice, leading to thermal expansion upon freezing.

@@

Several classical density functionals have already been developed for
water~\cite{ding1987, Yang1992, Jaqaman2004}, capturing some of the
qualitative behavior of water.  However, many of these are less
sophisticated than those describing the hard-sphere fluid, and a
quantitatively accurate functional describing water---which reproduces both
the macroscopic surface tension and its decrease as curvature
increases---has not yet been published.  For instance, Jaqaman \emph{et al}
developed a classical density-functional for water including orientational
effects, but their functional resulted in a surface tension twice the
experimental value, with the result that it can really only be used to make
very qualitative predictions~\cite{Jaqaman2004}.

@@

We use classical DFT to construct a new functional that will predict
the density of water molecules in a classical system.  The model we use
starts with a treatment of each individual water molecule as a purely
classical hard sphere and adds association terms based on Statistical 
Associating Fluid Theory (SAFT). Water, unlike simple
fluids which are mainly defined by van der Waals forces and weak electrostatic
interations, requires direct treatment of association effects. The strong 
hydrogen bonds and polar
interactions in water are ideal for describing with SAFT, which was designed for
use with such fluids. It is based on Wertheim's first-order
thermodynamic perturbation theory (TPT1)
\cite{wertheim1984fluidsI,wertheim1984fluidsII,wertheim1986fluidsIII,
wertheim1986fluidsIV} which provides a useful way to account for strong
associative interations between molecules. SAFT can account for the association energies
as well as any effects due to clusters or chains forming
in the liquid\cite{muller2001molecular}. \needsworknow{Expand or move this 
introduction to SAFT?}

\section{Theory and Methods}
We introduce a new classical density functional for water that
predicts the vapor pressure, vapor density, liquid density, bulk
surface tension, and enthalpy, as well as allows for liquid-vapor
coexistence.  The Helmholtz free energy functional is based on SAFT,
and is composed of the usual terms:
\begin{equation}
  F[n] = F_\textit{id}[n] + F_\textit{hs}[n]  +
F_\textit{disp}[n]+ F_\textit{assoc}[n],
\end{equation}
where $F_\textit{id}$ is the ideal gas free energy, $F_\textit{hs}$ is
the hard-sphere free energy, $F_\textit{disp}$ is the dispersion energy,
and $F_\textit{assoc}$ is the free energy of association.  In
addition, a chemical potential term is used, so we actually work in
the grand canonical ensemble.  In the following sections, we will
introduce the terms of this functional.

@@

\newcommand\etadisp{\ensuremath{\eta_\textit{d}}}
\newcommand\epsilondisp{\ensuremath{\epsilon_\textit{d}}}
\newcommand\epsilonassoc{\ensuremath{\epsilon_\textit{a}}}
\newcommand\kappaassoc{\ensuremath{\kappa_\textit{a}}}
\newcommand\lambdadisp{\ensuremath{\lambda_\textit{d}}}
\newcommand\lscale{\ensuremath{s_d}}
\subsection{Dispersion free energy}
The first term in the attractive energy is the dispersion term.  This
will include van der Waals attraction and any orientation-independent
interactions. We use a dispersion term based on the SAFT-VR
approach\cite{gil-villegas-1997-SAFT-VR}, which has two free
parameters, an interaction energy $\epsilondisp$ and a
length scale $\lambdadisp R$.

@@

A common model of hydrophobic solutes is the \emph{hard sphere solute}
\fixme{add citations?}.  The effect of this solute is merely to create
a cavity within which the water may not penetrate.  This is the
simples possible solute, and will serve as a test case, which if
passed could lead to the ability to insert different chemicals into
aqueous solution, instead of a vapor cavity, in order to observe
biologically and chemically important aqueous solvation interactions.

>>>>

Water stands out among liquids for its extraordinary properties.
Water is the only natural substance that is found in solid, liquid and
gaseous states at temperatures and pressures normally found on Earth.
Water has a density maximum at 4$^\circ$C, and expands upon freezing.
Water has the highest heat capacity, latent heat of fusion and
vaporization, thermal conductivity and surface tension of any
non-metallic liquid.  Water has an unusually high dielectric constant.
Much of the unusual behavior of water is explained by the ability of
each water molecule to form \emph{four} strong, directional hydrogen
bonds---two acceptors and two donors.  This directional bonding
dictates the open structure of ice, leading to thermal expansion upon
freezing.

@@

Several classical density functionals have already been developed for
water~\cite{ding1987, Yang1992, Jaqaman2004}, capturing some of the
qualitative behavior of water.  However, many of these are less
sophisticated than those describing the hard-sphere fluid, and a
quantitatively accurate functional describing water---which reproduces
both the macroscopic surface tension and its decrease as curvature
increases---has not yet been published.  For instance, Jaqaman
\emph{et al} developed a classical density-functional for water
including orientational effects, but their functional resulted in a
surface tension twice the experimental value, limiting it to purely
qualitative predictions~\cite{Jaqaman2004}.

@@

\section{Theory and Methods}
We use classical DFT to construct a new functional that will predict
the density of water molecules in a classical system.  The model we use
starts with a treatment of each individual water molecule as a purely
classical hard sphere and adds association terms based on Statistical 
Associating Fluid Theory (SAFT). Water, unlike simple
fluids which are mainly defined by van der Waals forces and weak electrostatic
interations, requires direct treatment of association effects. The strong 
hydrogen bonds and polar
interactions in water are ideal for describing with SAFT, which was designed for
use with hydrogen-bonding fluids. It is based on Wertheim's first-order
thermodynamic perturbation theory (TPT1)
\cite{wertheim1984fluidsI,wertheim1984fluidsII,wertheim1986fluidsIII,
wertheim1986fluidsIV} which provides a useful way to account for strong
associative interations between molecules. SAFT can account for the association energies
as well as any effects due to clusters or chains forming \fixme{not true}
in the liquid\cite{muller2001molecular}.  \needsworknow{Expand this a
  bit to accurately describe SAFT and its advantages.}

The Helmholtz free energy functional is based on SAFT,
and is composed of the terms:
\begin{equation}
  F[n] = F_\textit{id}[n] + F_\textit{hs}[n]  +
F_\textit{disp}[n]+ F_\textit{assoc}[n],
\end{equation}
where $F_\textit{id}$ is the ideal gas free energy, $F_\textit{hs}$ is
the hard-sphere free energy, $F_\textit{disp}$ is the dispersion energy,
and $F_\textit{assoc}$ is the free energy of association.  In
addition, a chemical potential term is used, so we actually work in
the grand canonical ensemble.  In the following sections, we will
introduce the terms of this functional.

@@

\newcommand\etadisp{\ensuremath{\eta_\textit{d}}}
\newcommand\epsilondisp{\ensuremath{\epsilon_\textit{d}}}
\newcommand\epsilonassoc{\ensuremath{\epsilon_\textit{a}}}
\newcommand\kappaassoc{\ensuremath{\kappa_\textit{a}}}
\newcommand\lambdadisp{\ensuremath{\lambda_\textit{d}}}
\newcommand\lscale{\ensuremath{s_d}}
\subsection{Dispersion free energy}
The first term in the attractive energy is the dispersion term.  This
will include van der Waals attraction and any orientation-independent
interactions. We use a dispersion term based on the SAFT-VR
approach\cite{gil-villegas-1997-SAFT-VR}, which has two free
parameters: an interaction energy $\epsilondisp$ and a
length scale $\lambdadisp R$.

@@

A common model of hydrophobic solutes is the \emph{hard sphere solute}
\fixme{add citations?}.  The effect of this solute is merely to create
a cavity within which the water may not penetrate.  This is the
simplest possible solute, and will serve as a test case, which if
passed could lead to the ability to insert different chemicals into
aqueous solution, instead of a vapor cavity, in order to observe
biologically and chemically important aqueous solvation interactions.
